---
title: "STAT 230A Final Project"
# author: "Andrej Leban"
# email: "andrej_leban@berkeley.edu"
# date: "`r format(Sys.time(), '%B %d, %Y')`"
header-includes:
  - \usepackage{float, amsthm, amsmath, amssymb, bm, enumitem, latexsym, color, minipage-marginpar, caption, multirow, verbatim}
output:
  pdf_document:
    toc: yes
    number_sections: true
papersize: a4
urlcolor: blue
# bibliography: [references.bib]
editor_options:
  markdown:
    wrap: 120
always_allow_html: true
---

```{r setup, echo = FALSE, message=FALSE, warning=FALSE}
# load in useful packages
library(tidyverse)
library(viridis)

# extra:
library(gsubfn)
library(reshape2)

library(data.table)

library(knitr)

library(foreign)
library(haven)

# set default knitr chunks
knitr::opts_chunk$set(
  echo = TRUE,
  fig.align = "center",
  fig.height = 4,
  fig.pos = "H",
  fig.width = 6,
  message = FALSE,
  warning = FALSE,
  cache = F,
  results = 'hide'
)

# define my_theme - append here
my_theme <- theme_minimal()
```

`TODO: margins, font size, group name & name in upper right`

<!-- LaTeX preamble -->
  \newcommand{\E}{{\mathbb E}}
\newcommand{\se}{{\mathcal{E}}}
\newcommand{\R}{{\mathbb R}}
\renewcommand{\P}{{\mathbb P}}
\newcommand{\ac}{{\mathcal{A}}}
\newcommand{\A}{{\mathcal{A}}}
\newcommand{\B}{{\mathcal{B}}}
\newcommand{\T}{{\mathcal{T}}}
\newcommand{\Z}{{\mathcal{Z}}}
\newcommand{\K}{{\mathcal{K}}}
\newcommand{\lc}{{\mathcal{L}}}
\newcommand{\Ps}{{\mathcal{P}}}
\newcommand{\pa}{{\mathring{p}}}
\newcommand{\F}{{\cal F}}
\newcommand{\samp}{{\mathcal{X}}}
\newcommand{\X}{{\mathcal{X}}}
\newcommand{\hi}{{\hat{\Phi}}}

\def\qt#1{\qquad\text{#1}}

\def\argmin{\mathop{\rm argmin}}
\def\argmax{\mathop{\rm argmax}}

\setlength{\parskip}{1.4 \medskipamount}


```{r, data_import, }
table3Df <- read_dta("data_raw/Table_3b.dta")

```


#  Paper Summary & Summary Statistics Table


# Main Result Replication


# Robustness check replication


# "Re-analysis"





